---
title: "Oil & Gas: Time-Series Forecasting"
author: "Emad Abdellatif"
date: "October 27th, 2024"
output:
  html_document:
    df_print: default
    highlight: tango
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
       message = FALSE,
       warning = FALSE)
```

# (1) Data Cleaning & Exploration


```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(forecast)
library(tseries)
library(modeltime)
library(tidymodels)
library(prophet)
library(timetk)


setwd("~/Desktop/Data Science Projects/Oil & Gas Forecasting")

data <- read.csv("well_data.csv")
```

## "series" dataframe

```{r}
series <- data %>%
  select(period, well_name) |>
  mutate(value = 1) %>% 
  pivot_wider(names_from = well_name, values_from = value, values_fill = list(value = 0)) %>%
  arrange(period)

series |>
  slice_head(n=20)

```
<br>

## Well Characteristics Data Frame

```{r}
well_characteristics <- data %>%
  rename(date = period) |>
  group_by(well_name) %>%
  summarise(
    Avg_Gas_to_Oil_Ratio = mean(gas_total / oil, na.rm = TRUE),
    Months_of_Production = n(),
    Initial_Production_Date = min(date),
    Avg_Monthly_Gas_Decline_Rate = {
      gas_diff <- diff(gas_total)
      gas_lag <- lag(gas_total, default = first(gas_total))[-1]  
      rate <- gas_diff / gas_lag
      rate[is.infinite(rate)] <- NA 
      mean(rate, na.rm = TRUE)
    }
  ) %>%
  ungroup()

print(well_characteristics)

```

<br>


# (2) Forecasting

## Filtering out Wells < 24 months and data with zero production

```{r}
wells_ge_24 <- well_characteristics |>
  filter(Months_of_Production >= 24)

cleaned_data <- data |>
  filter(well_name %in% wells_ge_24$well_name & oil != 0 & gas_total != 0) |>
  rename(date = period)
```

<br>

## Data Cleaning & Exploration


Combine data for all Wells into a single aggregated dataframe
```{r}
data_aggregated <- cleaned_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) |>
  group_by(date) %>%
  summarise(total_oil = sum(oil, na.rm = TRUE),
            total_gas = sum(gas_total, na.rm = TRUE)) |>
  arrange(date)

head(data_aggregated)

```
<br>

Plot the full time series to observe the behavior of the data.
```{r}
ggplot(data_aggregated, aes(x = date, y = total_oil)) +
  geom_line(color = "blue") + 
  labs(title = "Oil Production Over Time",
       x = "Month",
       y = "barrels/d") +
  theme_minimal()


```
<br>

The data appears to display high volatility between 1980 and 2010 with the trend showing stabilization after 2010. Due to this, I made the decision to remove the data prior to 2010 and focus on 2010-2018 data for forecasting. Given that the goal is to forecast the next 6 months, the last 8 years of data seems to be more applicable.

### 2010 - present Data
```{r}
data_subset <- data_aggregated |>
  filter(date >= as.Date("2010-01-01"))


data_subset_oil <- data_subset |>
  select(date, total_oil)

data_subset_oil |> plot_time_series(date, total_oil)

```

<br>

Looking at the plot above, we can see that there may be an outlier between 2015 and 2016. I'll go ahead and include a step for handling outliers.


### Handling Outliers
```{r}
Q1 <- quantile(data_subset_oil$total_oil, 0.25)
Q3 <- quantile(data_subset_oil$total_oil, 0.75)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- which(data_subset_oil$total_oil < lower_bound | data_subset_oil$total_oil > upper_bound)

median_value <- median(data_subset_oil$total_oil, na.rm = TRUE)
data_subset_oil$total_oil[outliers] <- median_value

```
<br>

Let's take a look at the data after handling outliers

```{r}
data_subset_oil |> plot_time_series(date, total_oil)

```

<br>


### Test for Stationarity

I'll use the ADF test to test for stationarity

```{r}
strtyr <- min(year(data_subset_oil$date))

ts_data <- ts(data_subset_oil$total_oil,
              frequency = 12,
              start = c(strtyr, 1))

adf_oil <- adf.test(ts_data, alternative = "stationary")
print(adf_oil)

```
<br>

With a p-value of 0.6743, the results show no-stationarity. I'll try first differencing.

### Differencing

```{r}

diff_ts_data <- diff(ts_data) 
adf_test_diff <- adf.test(diff_ts_data, alternative = "stationary")
print(adf_test_diff)

```

P-value is now below 0.05, showing stationarity after differencing.


<br>

I'll now go ahead and split the data into training and test sets and observe the performance of the models *auto_arima*, *prophet*, *prophet_xgboost*, *glmnet*, and *auto_arima_boost*.

## Model Testing

### Train/Test splits
```{r}

splits <- time_series_split(
  data_subset_oil,
  assess     = "12 months",
  cumulative = TRUE
)

splits %>%
  tk_time_series_cv_plan() %>%
  plot_time_series_cv_plan(date, total_oil)
```

<br>

### Model Fitting

```{r}
# FORECAST ----

# * AUTO ARIMA ----
model_arima <- arima_reg() %>%
  set_engine("auto_arima") %>%
  fit(total_oil ~ date, training(splits))

# * Prophet ----
model_prophet <- prophet_reg(
  seasonality_yearly = FALSE,
) %>%
  set_engine("prophet") %>%
  fit(total_oil ~ date, training(splits))

# * Prophet XGBoost----
model_prophet_xg <- prophet_boost(
  seasonality_yearly = FALSE
) %>%
  set_engine("prophet_xgboost") %>%
  fit(total_oil ~ date + as.numeric(date) + month(date, label = TRUE),
      training(splits))

# * Machine Learning - GLM ----
library(glmnet)
model_glmnet <- linear_reg(penalty = 0.1
                           ,mixture = 0.5) %>%
  set_engine("glmnet") %>%
  fit(
    total_oil ~ month(date, label = TRUE)
    + as.numeric(date),
    training(splits)
  )

# Auto ARIMA with XGBoost
model_arima_boost <- arima_boost() |>
  set_engine("auto_arima_xgboost") |>
  fit(total_oil ~ date + as.numeric(date) + month(date, label = TRUE),
      training(splits))
```

<br>

### Model Comparison

```{r}
model_tbl <- modeltime_table(
  model_arima,
  model_prophet,
  model_prophet_xg,
  model_glmnet,
  model_arima_boost
)

```
<br>

Calculate the accuracy of the models on the training set and the test set.
```{r}
calib_tbl_train <- model_tbl %>%
  modeltime_calibrate(training(splits))

calib_tbl <- model_tbl %>%
  modeltime_calibrate(testing(splits))

calib_tbl_train %>% modeltime_accuracy()
calib_tbl %>% modeltime_accuracy()
```

Based on the summaries of the models between both the training set and the test set, the model **ARIMA (0,1,1) with XGBoost** seems to be performing the best out of the 4 other models after reviewing the **rmse** and **mae** values between both the training set and the test set. 
<br>


### Plot Residuals

Plot the residuals of the training set
```{r}
residuals_timeplot_train <- model_tbl |>
  modeltime_calibrate(training(splits)) |>
  modeltime_residuals() |>
  plot_modeltime_residuals(
    .type = "timeplot",
    .interactive = TRUE
  )

residuals_timeplot_train
```

Looking at the residuals, the models **Prophet with XGBoost**, and **ARIMA(0,1,1) with XGBoost** seem to be capturing the overall trend and variability in the data reasonably well.

<br>


### Plot ACF

Plot the ACF of the residuals for the training set
```{r}
residuals_acf_train <- model_tbl |>
  modeltime_calibrate(training(splits)) |>
  modeltime_residuals() |>
  plot_modeltime_residuals(
    .type = "acf",
    .interactive = TRUE
  )

residuals_acf_train

```

**ARIMA(0,1,1)**, **Prophet with XBoost**, and **ARIMA(0,1,1) with XGBoost** show no significant auto-correlation, although **ARIMA (0,1,1) with XGBoost** crosses the bounds of the confidence interval slighlty more than **ARIMA (0,1,1)** and **Prophet with XGBoost**.
<br>

### Forecast on Original data

Display the forecast for the next 6 months using the 5 models that were tested, for comparison.
```{r}
future_forecast_tbl <- calib_tbl %>%
  modeltime_refit(data_subset_oil) %>%
  modeltime_forecast(
    h           = "6 months",
    actual_data = data_subset_oil
  ) %>%
  plot_modeltime_forecast()

future_forecast_tbl
```
<br>


In summary, I started my exploratory analysis by evaluating the long term trend of the oil production data and the benefit of having the full data during forecasting. I concluded that the data from 2010 and onwards seemed more suitable and applicable for forecasting the next 6 months.

Outliers were observed during the period between 2015 and 2016 that could've affected the modeling results moving forward. I went ahead and addressed the outliers using the IQR method for outlier detection.

In the forecasting phase, I utilized the following models **ARIMA**, **ARIMA with XGBoost**, **Prophet**, **Prophet with XGBoost**, and **Elastic Net (GLMNET)**. Based on the results of the models and the distribution of the residuals, it appears that **ARIMA with XGBoost** performs the best based on the *rmse* and *mae* along with the residual plots of both the training and the test sets.



# (3) Visualization

<p>
  <a class="btn btn-info btn-sm" data-toggle="collapse" data-target="#collapseToggle" role="button" aria-expanded="false" aria-controls="collapseExample">
    Show/Hide code
  </a>
</p>

<div class="collapse" id="collapseToggle">
  <div class="card card-body">
```{r}
insert_well_forecast <- function(well) {
  
  df <- cleaned_data |> 
    filter(well_name == well) |>
    mutate(Months_of_Production = row_number()) |>
    rename(total_oil = oil) |>
    mutate(date = as.Date(date, format = "%Y-%m-%d"))

  ### Handling Outliers
  Q1 <- quantile(df$total_oil, 0.25)
  Q3 <- quantile(df$total_oil, 0.75)
  IQR <- Q3 - Q1

  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR

  outliers <- which(df$total_oil < lower_bound | df$total_oil > upper_bound)

  median_value <- median(df$total_oil, na.rm = TRUE)
  df$total_oil[outliers] <- median_value
  
  split_index <- round(max(df$Months_of_Production) * 0.2)
  
  df <- df |>
    select(date, total_oil, well_name)
  
  
  splits <- time_series_split(
    df,
    assess     = split_index,
    cumulative = TRUE
  )
  
  # FORECAST ----
  # * Prophet ----
  model_prophet <- prophet_reg(
    seasonality_yearly = FALSE
  ) %>%
    set_engine("prophet") %>%
    fit(total_oil ~ date, training(splits))
  
  # * Prophet XGBoost----
  model_prophet_xg <- prophet_boost(
    seasonality_yearly = FALSE
  ) %>%
    set_engine("prophet_xgboost") %>%
    fit(total_oil ~ date + as.numeric(date) + month(date, label = TRUE),
        training(splits))
  
  # * AUTO ARIMA ----
  model_arima <- arima_reg() %>%
    set_engine("auto_arima") %>%
    fit(total_oil ~ date, training(splits))
  
  # STLM ETS
  model_arima_boost <- arima_boost() |>
    set_engine("auto_arima_xgboost") |>
    fit(total_oil ~ date + as.numeric(date) + month(date, label = TRUE),
        training(splits))
  
  model_arima_boost
  
  
  # MODELTIME COMPARE ----
  
  # * Modeltime Table ----
  model_tbl <- modeltime_table(
    model_arima_boost
  )
  
  
  # * Calibrate ----
  calib_tbl <- model_tbl %>%
    modeltime_calibrate(testing(splits))
  
  # * Accuracy ----
  model_accuracy <- calib_tbl %>% modeltime_accuracy()
  model_accuracy
  
  future_forecast_tbl <- calib_tbl %>%
    modeltime_refit(df) %>%
    modeltime_forecast(
      h           = "6 months",
      actual_data = df
    )
  
  plot_data <- future_forecast_tbl |>
    select(.model_desc, .key, .index, .value, .conf_lo, .conf_hi) |>
    rename(model = .model_desc,
           key = .key,
           date = .index,
           value = .value,
           conf_lo = .conf_lo,
           conf_hi = .conf_hi) |>
    mutate(Months_of_Production = row_number())
  
  
  display_forecast <- ggplot(plot_data, aes(x = Months_of_Production, y = value, color = key)) +
                        geom_line() +
                        geom_ribbon(aes(ymin = conf_lo, ymax = conf_hi), fill = "blue", alpha = 0.2) +
                        labs(title = paste0("Projected Oil Production Growth (",well,") from",
                                            year(min(plot_data$date)),"-", year(max(plot_data$date))), x = "Date", y = "Value") +
                        scale_x_continuous(breaks = seq(1, max(plot_data$Months_of_Production), by = 12)) +
                        theme_minimal()+
                        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y = "Barrels/d",
         x = "Months")
  
  display_forecast
  
  
  return(display_forecast)
  
}

```
  </div>
</div>


```{r}
forecasted_viz_11A <- insert_well_forecast("FIELD11A")
forecasted_viz_58 <- insert_well_forecast("FIELD58")


forecasted_viz_11A
forecasted_viz_58

```


